HO unification from object language to meta language

### 2024-02-26 paper/main.tex
wip

In the case of the OL unification variables \elpiIn{fuv} have no scope.
For example the statement of the instance \coqIn{forall_dec} contains the
term \coqIn{P x} where \coqIn{P} stands for a function of type
\coqIn{A ~$\to$~ Prop}. However our representation of \coqIn{P x} in the meta
language is \elpiIn{app[P, x]} where \elpiIn{P} is of type \elpiIn{ftm}.
In the terms of the ML unification variables come with a scope.

\begin{elpicode}
  type uva nat -> list tm -> tm.
\end{elpicode}

\noindent
Of course some attention
should be payed when dealing with \elpiIn{uva}, since we have to certify
that an \elpiIn{uva i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is always a list of distinct names.

@@

\begin{elpicode}
  typeabbrev fsubst list (option fm).
\end{elpicode}

@@

\begin{elpicode}
  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
  typeabbrev subst list (option tm).
\end{elpicode}

@@

\subsection{Auxiliary relations}

@@

\begin{elpicodetab}
  type fderef fsubst -> fm -> fm -> o.
  fderef S (fuva N) R :- assigned? N S T, fderef S T R.
  fderef S (app [fuva N|L]) R :- 
    assigned? N S (app NL), append NL L AL, fderef S (app AL) R.
  fderef _ (fuva N) (fuva N).
  fderef S (fapp L1) (fapp L2) :- forall2 (fderef S) L1 L2.
  fderef S (flam F1) (flam F2) :- 
    pi x\ fderef S x x => fderef S (F1 x) (F2 x).
  fderef _ (fcon X) (fcon X).
\end{elpicodetab}

@@

\begin{elpicode}
  type deref subst -> tm -> tm -> o.
  deref S (uva N L) R :- assigned? N S T, beta-abs T L T', deref S T' R.
\end{elpicode}

In particular, 1) if the variable \elpiIn{fov N} is assigned to a term
\elpiIn{T} in a substitution mapping \elpiIn{S}, we return the result of
dereferencing \elpiIn{T}; 2) if the current term is the application of a
variable \elpiIn{fov N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
is assigned to a term \elpiIn{T}, we return the result of dereferencing the beta
reduction of \elpiIn{T} to \elpiIn{L}. This is due to the semantics of the
unification variables within the OL; 3) finally, if \elpiIn{uv N} is not
assigned, then the same variable is returned.

@@

\begin{elpicode}
  ho_deref H (hov N L) R :- assigned? N H T, 
    ho_beta T L R', ho_deref H R' R.
  ho_deref H (hov X L) (hov X L1) :- 
    forall2 (ho_deref H) L L1.
\end{elpicode}

\noindent
where, 1) if the variable \elpiIn{hov N L} has an assignment \elpiIn{T}, we
return the dereferencing of the beta reduction \elpiIn{T} to the scope
\elpiIn{L}.

\begin{figure}
  \begin{elpicode}
    type ~(\Eo)~ tm -> tm -> o.
    % congruence
    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
    con X ~\Eo~con X.
    fov N ~\Eo~fov N.
    % eta
    lam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
      x ~\Eo~x => F x ~\Eo~T' x.
    T ~\Eo~lam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
      x ~\Eo~x => T' x ~\Eo~F x.
    % beta
    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
  \end{elpicode}
  \caption{Equal predicate}
  \label{code:eq-pred}
  \Description[code:equal-pred]{code:equal-pred}
\end{figure}

@@

\begin{figure}
  \begin{elpicode}
    type unif tm -> tm -> ho_subst -> ho_subst ->  o.
    % Base cases
    unif (app L1) (app L2) S S1 :- fold2 unif L1 L2 S S1.
    unif (lam F1) (lam F2) S S1 :- 
      pi x\ (pi S\ unif x x S S) => unif (F1 x) (F2 x) S S1.
    unif (con X) (con X) S S.
    % Assigned variables
    unif (hov N Args) T S S1 :- 
      assigned-deref N S F, beta-abs F Args T1, unif T1 T S S1.
    unif T (hov N Args) S S1 :- 
      assigned-deref N S F, beta-abs F Args T1, unif T T1 S S1.
    % Non-assigned variabless
    unif (hov N Args1) (hov M Args2) S S3 :-  
      pattern-fragment Args1, pattern-fragment Args2, 
      new S W S1, prune W Args1 Args2 Ass, 
      assign N S1 Ass S2, assign M S2 Ass S3.
    unif (hov N Args) T S S1 :- pattern-fragment Args,
      not (T = hov _ _), not_occ N S T,  
      bind Args T T1, assign N S T1 S1.
    unif T (hov N Args) S S1 :- pattern-fragment Args,
      not (T = hov _ _), not_occ N S T, 
       bind Args T T1, assign N S T1 S1.
  \end{elpicode}
  \caption{The \Ue predicate}
  \label{code:ue-pred}
  \Description[code:unif-pred]{code:unif-pred}
\end{figure}

>>>>

In the case of the OL unification variables \elpiIn{fuv} have no scope.
For example the statement of the instance \coqIn{forall_dec} contains the
term \coqIn{P x} where \coqIn{P} stands for a function of type
\coqIn{A ~$\to$~ Prop}. However our representation of \coqIn{P x} in the meta
language is \elpiIn{app[P, x]} where \elpiIn{P} is of type \elpiIn{ftm}.
In the terms of the ML unification variables come with a scope (the second
argument of the \elpiIn{uva} constructor).
Of course some attention
should be payed when dealing with \elpiIn{uva}, since we have to certify
that an \elpiIn{uva i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is always a list of distinct names.

@@

\marginpar{maybe move code to the appendix}
\begin{elpicode}
typeabbrev memory A (list (option A)).
type set nat -> memory A -> A -> o.
set N S T :- nth N S (some T).
type unset nat -> memory A -> o.
unset N S :- nth N S none.
type assign nat -> memory A -> A -> memory A -> o.
assign z [none|M] T [some T|M].
assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
\end{elpicode}

@@

\begin{elpicode}
  typeabbrev fsubst (memory ftm).

  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
  typeabbrev subst (memory assmt).
\end{elpicode}

@@

\subsection{equational theory and unification}

here we give the functions/signatures to express the properties 3-5 in the problem
statement

@@

\begin{elpicode}
type fderef fsubst -> fm -> fm -> o.
fderef S (fapp [fuva N|A]) R :- set N S (fapp B),
  append A B C, fderef S (fapp C) R.
fderef S (fapp [fuva N|A]) (fapp [fuva N|B]) :- unset N S,
  forall2 (fderef S) A B.
fderef S (fapp A) (fapp B) :- forall2 (fderef S) A B.
fderef S (flam F) (flam G) :- 
  pi x\ fderef S x x => fderef S (F x) (G x).
fderef _ (fcon C) (fcon C).
fderef S (fuva N) R :- set N S T, fderef S T R.
fderef _ (fuva N) (fuva N) :- unset N S.
\end{elpicode}

@@

\marginpar{what about app uva?}
\begin{elpicode}
  type deref subst -> tm -> tm -> o.
  % ... the same
  deref S (uva N A) R :- set N S T,
    instantiate T A T', deref S T' R.
  deref S (uva N A) (uva N B) :-unset N S,
    forall2 (deref S) A B.
\end{elpicode}

In particular, 1) if the variable \elpiIn{fuva N} is assigned to a term
\elpiIn{T} in a substitution mapping \elpiIn{S}, we return the result of
dereferencing \elpiIn{T}; 2) if the current term is the application of a
variable \elpiIn{fuva N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
is assigned to a term \elpiIn{T}, we return the result of dereferencing the beta
reduction of \elpiIn{T} to \elpiIn{L}. This is due to the semantics of the
unification variables within the OL; 3) finally, if \elpiIn{uv N} is not
assigned, then the same variable is returned.

@@

where, 1) if the variable \elpiIn{hov N L} has an assignment \elpiIn{T}, we
return the dereferencing of the beta reduction \elpiIn{T} to the scope
\elpiIn{L}.

\begin{figure}
  \begin{elpicode}
    type ~(\Eo)~ ftm -> ftm -> o.
    % congruence
    fapp A ~\Eo~fapp B :- forall2 ~(\Eo)~ A B.
    flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
    fcon C ~\Eo~fcon C.
    fuva N ~\Eo~fuva N.
    % eta
    flam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
      x ~\Eo~x => F x ~\Eo~T' x.
    T ~\Eo~flam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
      x ~\Eo~x => T' x ~\Eo~F x.
    % beta
    fapp [flam X | TL] ~\Eo~T :- beta (flam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
    T ~\Eo~fapp [flam X | TL] :- beta (flam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
  \end{elpicode}
  \caption{Equal predicate OL}
  \label{code:eq-pred}
  \Description[code:equal-pred]{code:equal-pred}
\end{figure}

\begin{figure}
  \begin{elpicode}
    type ~(\Ee)~ tm -> tm -> o.
    app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
    lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
    con C ~\Ee~fcon C.
    uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
  \end{elpicode}
  \caption{Equal predicate ML}
  \label{code:eq-pred-meta}
  \Description[code:equal-pred-meta]{code:equal-pred-meta}
\end{figure}

@@

\begin{figure}
\begin{elpicode}
type ~(\Ue)~ tm -> tm -> subst -> subst ->  o.
% Congruence
(app A ~\Ue~app B) S S1 :- fold2 ~\Ue~A B S S1.
(lam F ~\Ue~lam G) S S1 :- 
  pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) S S1.
~\Ue~(con C) (con C) S S.
% deref
(uva N A ~\Ue~T) S S1 :- set N S F,
  beta-abs F A T1, (T1 ~\Ue~T) S S1.
(T ~\Ue~uva N A) S S1 :- set N S F,
  beta-abs F A T1, (T ~\Ue~T1) S S1.
% flex-flex
(uva N A ~\Ue~uva M B) S S3 :-  unset M, unset N,
  pattern-fragment A, pattern-fragment B, 
  new S W S1, prune W Args1 B Ass, 
  assign N S1 Ass S2, assign M S2 Ass S3.
% assignment
(uva N A ~\Ue~T) S S1 :- pattern-fragment A,
  not (T = uva _ _), not_occ N S T,  
  bind A T T1, assign N S T1 S1.
(T ~\Ue~uva N A) S S1 :- pattern-fragment A,
  not (T = uva _ _), not_occ N S T, 
    bind A T T1, assign N S T1 S1.
%  Note: (a infix b) c d === (infix) a b c d

\end{elpicode}
\caption{The \Ue predicate // TODO move to the appendix}
\label{code:ue-pred}
\Description[code:unif-pred]{code:unif-pred}
\end{figure}
